---
# UNL thesis fields
title: "DATA SCIENCE, DASHBOARDS, AND THE WAY IT WORKS WITH STATISTICS"
author: "Denise Renee Bradford"
month: "Month"
year: "Year"
location: "Lincoln, Nebraska"
major: "Statistics"
adviser: "Susan R. VanderPlas, Ph.D"
abstract: |
  Here is my abstract. *(350 word limit)*
acknowledgments: |
  Thank you to all my people!
dedication: |
  Dedicated to...
# End of UNL thesis fields
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    pandoc_args: --top-level-division=chapter
    keep_tex: yes
    latex_engine: xelatex
    template: template.tex
  huskydown::thesis_gitbook: 
    style: style.css
#  huskydown::thesis_word: default
#  huskydown::thesis_epub: default
bibliography: bib/thesis.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: bib/apa.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
#header-includes:
#- \usepackage{tikz}
---

<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of metadata used to produce the document.  Be careful with spacing in this header!

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.
-->

<!--
If you receive a duplicate label error after knitting, make sure to delete the index.Rmd file and then knit again.
-->

```{r setup, include = F}
options(width = 60)
knitr::opts_chunk$set(
  fig.width = 6, fig.height = 4, 
  out.width = "\\linewidth", dpi = 300, 
  tidy = T, tidy.opts=list(width.cutoff=45)
)
```

```{r include_packages, include = FALSE}
# This chunk ensures that the huskydown package is
# installed and loaded. This huskydown package includes
# the template files for the thesis.
if(!require(devtools))
  install.packages("devtools", 
                   repos = "http://cran.rstudio.com")
# if(!require(huskydown))
#   devtools::install_github(
#     "benmarwick/huskydown"
#   )
#library(huskydown)
library(knitr)
library(palmerpenguins)
library(tidyverse)
```

<!-- The {.unnumbered} option here means that the introduction will be "Chapter 0." You can also use {-} for no numbers on chapters. -->

<!-- # Introduction {.unnumbered} -->

# Literature Review

## Overview

Data Visualizations can be described as a graphical representation of tabular data and information. Data visualization tools are used in ways to accessible way to see and understand trends, outlines, and patterns in data. 



A list of Data Visualizations Types can be broken down into the following:

- Tables (Pivot Tables)
- Line Chart or Line Graph
- Bar Chart
- Scatter Plot
- Pie Chart or Doughnut Chart
- Trees/Hierarchical Maps
- Network
- Heat Maps

## Applications of Data Visualization Types

### Tables
A table is defined as set of objects or individuals divided by intersecting components, which produces a data table or tabular data.  A table will allow for a reader to perceive information at a high level in its totality.

```{r include=FALSE, fig.align="center"}
data("penguins")
table(penguins$species, penguins$sex)
```


### Line Chart (Line Graphs)

```{r include=FALSE, fig.align="center"}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) + geom_line()
```


### Bar Chart

```{r include=FALSE, fig.align="center"}
ggplot(penguins, aes(x=species, y=body_mass_g)) + 
  geom_bar(stat = "identity")
```



### Scatter Plot

```{r include=FALSE, fig.align="center"}
ggplot(penguins, aes(x=bill_length_mm, y=bill_depth_mm, color=species)) + 
    geom_point(size=4) 
```


### Pie Chart or Doughnut Chart

```{r include=FALSE, fig.align="center"}
ggplot(penguins, aes(x="", y=body_mass_g, fill=species)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
```


### Trees/Hierarchical Maps

```{r include=FALSE, fig.align="center"}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) + geom_line()
```


### Network

```{r include=FALSE, fig.align="center"}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) + geom_line()
```



### Heat Maps

```{r include=FALSE, fig.align="center"}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) + geom_line()
```



Visualizations of data are important for exploratory data analysis (EDA) along with model diagnostics. Plots for EDA are a useful tool for guiding an analyst in discovering the relationships between variables in their data. In the case when using plots in model diagnostics, plots help analysts determine whether or not the model is appropriate way to model. During the initial EDA stage, an analyst may find that a variable or a covariate is directly related to the dependent variable when looking at a correlation heatmap or a scatterplot. This will be important to know before starting a linear model analysis. Much of our general understanding is from introductory statistics courses. The basic understanding can be formalized in a way that will visual discovery process. 

Visualizations have become more effective in more recent years due to the pandemic and the Johns Hopkins University COVID-19 Dashboard [Dashboard](https://coronavirus.jhu.edu/map.html), for most of the world, we were glued to our computers, TVs and phones. As a result, we watched as the dashboard changed in real time to adapt to the needs of the users. In part to data growing and changing, the dashboard as well as the visualizations were needed in a condense platform. The need to be concise along with vastly informative can be a bit of a struggle when it comes to data visualizations. The human brain is capable of only taking in a set amount of data at a time from a table or a paragraph. The space of infographics have been a much better way of looking at data on a creative scale. While this may be a way of seeing the data in a nice way, infographics miss the interactive piece about data that many people would like to explore. 

## Visual Inference

Visual inference uses our ability to detect graphical anomalies. The idea of formal testing remains the same in visual inference – with one exception: The test statistic is now a graphical display which is compared to a “reference distribution” of plots showing the null.[REFERENCE WEBSITE]


### Applications

Two general applications in areas of visual inference have developed since the work of [@Buja (2009)]. These applications are actual methodology and methodology based on protocols.  Actual methodology applications are used with alternative and corresponding null hypotheses, which perform visual inference tests to show many participants of different backgrounds with lineups. 


## Parallel Coordinate Plot Visualization 

Parallel coordinate plots have been implemented in analysis software since the mid 1980s ([@Inselberg], [@Wegman]). Parallel Coordinate Plot Visualization, also called parallel plot, is an established field of high dimensional visualization of xy coordinate analysis. Parallel coordinate plots have such a unique structure that will allow for data structures with n-dimensions. As data grows in exploratory data analysis, the use of data visualizations will be more important and useful to understanding the underlying structures of useful data for analyses. Several packages in R are dedicated to visualizing parallel coordinate plots.

Dashboards can be used to help understand and support many types of data in any set of business objectives that are important. There are many different ways to label and utilize dashboards into different types. 

Dashboards are cognitive tools that should be used to improve understanding of data, which should help people visually find relationships, trends, patterns and outliers. Most importantly, dashboards should leverage people's visual cognitive capabilities. 


### R Packages

There are R packages that exist for Parallel coordinate plots, with some built-in functionality, but not many exist for advance exploratory data analysis visualizations. Beginning with base plot package `parcoord` developed in MASS by [@Venables and Ripley 2004]. The package is limited by extensive, in which it contains more than methods for the parallel coordinate plot visualization. It contains the following tools for 2D visualization of PCPs. The 2D layout contains a number of features that are [LIST FEATURES FROM PACKAGE HERE!!](). Another package `gclus` was developed by [@Hurley, 20219]() implements `cparcoord` to include panel color of the strength of a correlation between neighboring axes.

While base R packages are useful, there are a few packages within the ggplot2 environment implementing parallel coordinate plots. With ggplot2 extension packages, the theory of grammar of graphics are at the focal point of plots. The following packages [@GGally] and `ggpcp` are built with the fundamental guidelines for graphic literacy for numeric variables. `GGally` was originally designed for the following pieces of the graphic design: [LIST THE PACKAGE'S STARTING PLANS](), so it also contains much more capabilities for PCPs with addition to visualizations. `ggpcp` we separate the data transformations from the visualization, i.e. rather than working with a single function to draw a plot, we are providing a set of functions that work together [@ggpcp]().

`ggparallel` implements and combines different types of parallel coordinate plots for categorical data: hammock plots [@Schonlau 2003](), parallel sets plots, common angle plots, and common angle plots with a hammock-like adjustment for line widths [@ggparallel]().


### Why `gg` and `ggplot` are important?

The grammar of graphics (gg of ggplot2) is a theory that is well-defined for creating statistical graphics with work from [@Wilkinson (1999)] and [@Wickham (2009)]. The theory of graphics are based on plot layers, which is built with four distinct pieces: 

- the data
- aesthetic mapping - *takes data and maps the variable in the data frame to a particular visual features*
- statistical transformation - *determines how to transform the data to values to create the visual feature*
-  geometric object and a position adjustment - *geom object is used to draw a plot layer and position adjustment is to help with adjusting the visual feature in the space.*

### `ggpcp` package importance


## Summary


## Dashboards
### Formal Definition

Formally, a dashboard is a visual display of the most important information needed to achieve one or more objectives; consolidated and arranged on a single screen so the information can be monitored at a glance.[@few]. As dashboard has particular characteristics:

- Achieve specific objectives
- Fits on a single computer screen
- Information can be displayed in multiple mediums (web browser or mobile device)
- Can be used to monitor information at a high level

While a dashboard can be extremely useful, it may be worth describing that a poorly designed dashboard will not be used. A dashboard should be concise, clear, and intuitive when displaying components in combination of with a customized list of requirements of users.

Much of the work done within statistical research and dashboard design has to do with collaboration with other academic researchers. While this may be the best for the growth of the discipline, one will find that working with the 

A collaboration is defined when 2 or more entities work together to produce a desired and shared outcome.[@collab]() Interdisciplinary research with collaboration is a pinnacle aspect of dashboards and statistical graphics to produce innovation and scientific knowledge.

The National Academies defines research collaboration as follows [@National Academics Ref]():

> Interdisciplinary research (IDR) is a mode of research by teams or individuals that integrates information, data, techniques, tools, perspectives, concepts, and/or theories from two or more disciplines or bodies of specialized knowledge to advance fundamental understanding or to solve problems whose solutions are beyond the scope of a single discipline or field of research practice.


