---
# UNL thesis fields
title: "DATA SCIENCE, DASHBOARDS, AND THE WAY IT WORKS WITH STATISTICS"
author: "Denise Renee Bradford"
month: "Month"
year: "Year"
location: "Lincoln, Nebraska"
major: "Statistics"
adviser: "Susan R. VanderPlas, Ph.D"
abstract: |
  Here is my abstract. *(350 word limit)*
acknowledgments: |
  Thank you to all my people!
dedication: |
  Dedicated to...
# End of UNL thesis fields
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    pandoc_args: --top-level-division=chapter
    keep_tex: yes
    latex_engine: xelatex
    template: template.tex
  bookdown::gitbook:
    config:
      toc:
        collapse: subsection
        scroll_highlight: true
        before: null
        after: null
      toolbar:
        position: fixed
      edit : null
      download: null
      search:
        engine: lunr # or fuse
        # options to control/tune search engine behavior (for
        # fuse.js, refer to https://fusejs.io/api/options.html)
        options: null
      fontsettings:
        theme: white
        family: sans
        size: 2
      info: true
bibliography: bib/thesis.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: bib/apa.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
---

# UNL thesis fields

Placeholder



<!--chapter:end:index.Rmd-->


# Introduction

Placeholder


## Audience Considerations
### Perception
### Attention and Memory
### Constructing Meaning
### Expertise
### Engagement with the data
## Data Considerations
#### Interactive Graphics
## Audience-Data Interactions
### Testing static graphics
### Testing interactive graphics
## Dashboard Design
## Conclusion

<!--chapter:end:01-lit-review.Rmd-->


# Chapter Paper on Rural Shrink Smart Manuscript submitted to Journal of Data Science Special Issue {#rmd-basics}

Placeholder


## Abstract
## Introduction 
## Data Description
## Dashboard Design Considerations
## Guiding Design Principles
## Dashboard Design Process
### Dashboard Components 
### Initial Draft
### Redesign 
## Discussion
## Future Work 
## Conclusions

<!--chapter:end:02-chap1.Rmd-->

# Chapter 2 Stuff {#math-sci}

<!-- Required to number equations in HTML files -->
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>


## Many useful visualizations don't have easy interactivity

<!--chapter:end:03-chap2.Rmd-->

# Tables, Graphics, References, and Labels {#ref-labels}

## Dashboard EDA

<!--chapter:end:04-chap3.Rmd-->

# Conclusion {-}

If we don't want Conclusion to have a chapter number next to it, we can add the `{-}` attribute.

**More info**

And here's some other random info: the first paragraph after a chapter title or section head _shouldn't be_ indented, because indents are to tell the reader that you're starting a new paragraph. Since that's obvious after a chapter or section title, proper typesetting doesn't add an indent there.


<!--chapter:end:05-conclusion.Rmd-->

`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

<!--
If you feel it necessary to include an appendix, it goes here.
-->


# The First Appendix

This first appendix includes all of the R chunks of code that were hidden throughout the document (using the `include = FALSE` chunk tag) to help with readibility and/or setup.

**In the main Rmd file**

```{r ref.label='include_packages', results='hide', echo = TRUE}
```

**In Chapter \@ref(ref-labels):**

```{r ref.label='include_packages_2', results='hide', echo = TRUE}
```

# The Second Appendix, for Fun

<!--chapter:end:06-appendix.Rmd-->

# Colophon {-}

This document is set in [EB Garamond](https://github.com/georgd/EB-Garamond), [Source Code Pro](https://github.com/adobe-fonts/source-code-pro/) and [Lato](http://www.latofonts.com/lato-free-fonts/). The body text is set at 11pt with $\familydefault$. 

It was written in R Markdown and $\LaTeX$, and rendered into PDF using [huskydown](https://github.com/benmarwick/huskydown) and [bookdown](https://github.com/rstudio/bookdown). 

This document was typeset using the XeTeX typesetting system, and the [University of Washington Thesis class](http://staff.washington.edu/fox/tex/) class created by Jim Fox. Under the hood, the [University of Washington Thesis LaTeX template](https://github.com/UWIT-IAM/UWThesis) is used to ensure that documents conform precisely to submission standards. Other elements of the document formatting source code have been taken from the [Latex, Knitr, and RMarkdown templates for UC Berkeley's graduate thesis](https://github.com/stevenpollack/ucbthesis), and [Dissertate: a LaTeX dissertation template to support the production and typesetting of a PhD dissertation at Harvard, Princeton, and NYU](https://github.com/suchow/Dissertate)

The source files for this thesis, along with all the data files, have been organised into an R package, xxx, which is available at https://github.com/xxx/xxx. A hard copy of the thesis can be found in the University of Washington library.

This version of the thesis was generated on `r Sys.time()`. The repository is currently at this commit:

```{r echo = FALSE, eval = FALSE}
# I've set eval=FALSE to ensure Travis-CI can run
# if you're not using Travis-CI, then eval=TRUE will be fine
library(git2r)
if ( in_repository() ) {
       summary(commits()[[1]]) 
} else { 
       message("We are not in a git repository") 
}
```

The computational environment that was used to generate this version is as follows:

```{r echo = FALSE}
devtools::session_info()
```


<!--chapter:end:98-colophon.Rmd-->


# References {-}

Placeholder



<!--chapter:end:99-references.Rmd-->


# Testing Graphics 

Placeholder


### Variable Types
### Multivariate Data Displays
## `ggpcp` package - Generalized Parallel Coordinate Plots
## Parallel Coordinate Plot Visualization Review
### Parallel Coordinates in EDA
### R Packages
### `ggpcp` package importance

<!--chapter:end:graveyard.Rmd-->

---
output:
  pdf_document: default
---

# CONTEXT AND BACKGROUND

Topic: The relationship between good statistical graphics and appropriate dashboard design. 

**What is the current situation?**

- Data Scientists and Data Analysts are responsible for making decisions using large data sources for organizations. Large Data sources in decision-making have become more of the norm, which means that a consistent choice in dashboard design helps in exploratory data analysis (EDA). Limited resources and knowledge about the impact of large data sources can contribute to confusion and misconceptions about using large data in decision-making.    


**What is wrong with the situation and/or what needs to change?** 

- Many Dashboards are built to assist in data exploration, but many dashboards lack guidelines and appropriate perception with static and interactive statistical graphics. 


**What are the most recent events that led us to the current situation?**



**What is the setting of the problem?**

- Dashboards with statistical graphics


# KNOWLEDGE GAP


**What have researchers in similar settings explored or discovered?**



**What have researchers in other settings explored or discovered?** 



**What remains unknown about the problem?** 


- While it is known that interactive graphics is an essential aspect of statistical graphics and are used in dashboard designs frequently, the inconsistency gap between interactive graphics and dashboard design, it is unknown whether there is a relationship between interactive graphics and effective dashboard design can be implemented in Exploratory Data Analysis (EDA).

# SIGNIFICANCE


**Why is it important to fill the knowledge gap?**

- All dashboards will benefit from a comprehensive guidelines for effective dashboard design.


**Scope: How big is the problem, and how many people are affected in the field or specific context you are studying?**

- 

**Consequences: What are the consequences of not addressing the problem?**

- Dashboards will continue to be designed without the appropriate perception of good statistical interactive graphics when analysts develop dashboards for exploring large data. 

**Contribution: How will your study contribute to the problem’s solution? How does filling the gap help?**



# Draft Problem Statement




<!--chapter:end:ProblemStatementOutline.Rmd-->

